---
title: "MBIO 612 Final Project - Grain Size Distribution - Wet vs. Dry Seasons"
author: "Kyle Bosworth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.path = "../outputs/")
knitr::opts_chunk$set(dev = "png", 
                      fig.path = "../outputs/",
                      fig.width = 7,
                      fig.height = 5)
```

### Load Libraries
```{r}
library(here)
library(tidyverse)
library(leaflet)
library(tidyr)

```

### Read in Data
```{r}
#Had to manually load grainsize data. For some reason I could not get R to read in my csv file.


grainsize <- read.csv(here("MBIO612_finalproject","data", "grainsize.csv"))

head(grainsize)

#Had a lot of trouble with my naming schemes in my data frame, here i just wanted to double check my names
str(grainsize$SiteName)  # Should show character vector with names as MW_B5 not MW-B5

#Checked a few example site names too
head(unique(grainsize$SiteName))

```

### The Goal
I just wanted to focus on 2 sampling dates that occured in the dry and wet season, 9/14/23 and 12/20/23, my goald is to create map that will display grain size % and WQ data in an interactive form. Iʻd like to me able to make a story map of sorts and each site will have a pop up window that displays info for that day. 


### Data Prep
```{r}
#I want to first clean and prepare my data to remove NAs, and focus on spcific dates

grainsize_filter <- grainsize %>%
  mutate(
    Long = as.numeric(gsub("_", "-", as.character(Long))),
    Lat = as.numeric(as.character(Lat)) #this converts my long/lat values to numeric to makes sure that they are read in coordinate format. I replaced my "hypnons "-" with underscores and didnʻt want to go back and edi out each cell.
  ) %>%
  filter(!is.na(Long), !is.na(Lat)) %>%
  filter(Date %in% c("9/14/23", "12/20/23")) %>% # Remove rows where coordinates are NA and selecting for dates. 
  mutate(
    Gravel_pct = (Gravel/TotalWeight) * 100,
    CoarseSand_pct = (CoarseSand/TotalWeight) * 100,
    MediumSand_pct = (MediumSand/TotalWeight) * 100,
    FineSand_pct = (FineSand/TotalWeight) * 100,
    SiltClay_pct = (SiltClay/TotalWeight) * 100
  )
#here i calculated my grain size % based on total weight so that in my map i can view this as a percentage.


#I then thought that it might be easier to view all this dat if I made 2 different map. So i created 2 distinct seasonal data sets.

dry_season_data <- grainsize_filter %>% 
  filter(Date == "9/14/23")

wet_season_data <- grainsize_filter %>% 
  filter(Date == "12/20/23")
```


### Creating a pop-up
```{r}

#I would like to depict the sampling points on an interactive map that has a window that pops up when you click on a specific site. 

#I did some researching and found that I could make a pop-up using "popup_content" and structure the text using using html tags!
#the basics are as followed 


#"<strong>" wraps the lables 
#"<br/>" is a break and is used to create the next line
#round was used to round my percents, coordinates and data off to a degree.



#Line 1: "<strong>Environmental Parameters:</strong><br/>",
#Line 2: "<strong>Environmental Parameters:</strong><br/>",


grainsize_filter <- grainsize_filter %>%
  mutate(
    popup_content = paste(
      "<strong>Site:</strong>", SiteName, "<br/>",
      "<strong>Zone:</strong>", Zone, "<br/>",
      "<strong>Coordinates:</strong><br/>",
      "Lat: ", round(as.numeric(Lat), 6), "<br/>",
      "Long: ", round(as.numeric(Long), 6), "<br/>",
      "<strong>Grain Size Distribution:</strong><br/>",
      "Gravel: ", round(as.numeric(Gravel_pct), 1), "%<br/>",
      "Coarse Sand: ", round(as.numeric(CoarseSand_pct), 1), "%<br/>",
      "Medium Sand: ", round(as.numeric(MediumSand_pct), 1), "%<br/>",
      "Fine Sand: ", round(as.numeric(FineSand_pct), 1), "%<br/>",
      "Silt/Clay: ", round(as.numeric(SiltClay_pct), 1), "%<br/>",
      "<strong>Environmental Parameters:</strong><br/>",
      "Temperature: ", round(as.numeric(TempC), 2), "°C<br/>",
      "Salinity: ", round(as.numeric(Salinity), 2), "<br/>",
      "DO.: ", round(as.numeric(`DO.`), 2)
    )
  )
```

### Adding in the map + Heʻeia Fishpond Coordinates
```{r}

heeia_lat <- 21.4351  # This is the coordinates i used for my map layer
heeia_lng <- -157.8060

heeiamap <- function(data, season_name) {
  heeiamap_data <- data %>%
    filter(!is.na(Long), !is.na(Lat),
           Long < 0,  #makes sure that my longitude is negative (for Hawaiʻi), chk for me as my long had underscores
           Long > -158, # Set bounds
           Lat > 21,    
           Lat < 22)    
  
  #I wanted to define manual colors for mullet zones
  zone_colors <- c(
    "Mullet East" = "gold",
    "Mullet West" = "darkgreen",
    "Mullet South" = "royalblue",
    "Mullet North" = "red"
  )
  
  #i created a color palette function with the colors
  pal <- colorFactor(palette = zone_colors, domain = heeiamap_data$Zone)
  
  #Using the leaflet package i made my maps! creates a map using my heeiamap_data
  leaflet(heeiamap_data) %>%
    addTiles() %>%
    setView(lng = -157.8060, lat = 21.4351, zoom = 16) %>% #same as bounds and cenyers it on my heʻeia coordinates
    
    #here i added in circle markers to represent my sample sites on the maps
    addCircleMarkers(
    lng = ~Long,      #longitude from data
    lat = ~Lat,       #latitude from data
    popup = ~popup_content,  #brings up popup only when clicked
    color = ~pal(Zone),     #uses my color pal for zones
    radius = 8,            #changes the size of circle markers  on map
    fillOpacity = 0.7,     #70% opacity for circle markers
    stroke = TRUE,         #added an outer border
    weight = 1            #and a thickness to boarder
) %>%
    addLegend(
      position = "bottomright",
      pal = pal,  # Use the same custom palette
      values = ~Zone,
      title = paste(season_name, "Mullet Sampling Sites"),
      opacity = 0.7
    )
}

# Create seasonal datasets
dry_season_data <- grainsize_filter %>% 
  filter(Date == "9/14/23")
wet_season_data <- grainsize_filter %>% 
  filter(Date == "12/20/23")

# Generate and display maps
dry_map <- heeiamap(dry_season_data, "Dry Season")
wet_map <- heeiamap(wet_season_data, "Wet Season")

# Display maps
dry_map
wet_map


```

